### 2016-12-01 paper/chapters/vae.tex
add paragraph 'latent space' and 'formal setup'

\section{Variational Autoencoders}
\label{sec:vae}
Variational Autoencoders (VAE) have been introduced in 2014 by Kingma and Welling which uses variational inference to perform efficient inference and learning in deep latent models.
% maybe extra section/sub?
We assume that the high-dimensional input data $x$ can be explained using low-dimensional latent variables $z$.\\
While this might be surprising at first, most high-dimensional data we encounter can be explained using far fewer dimensions.
Let us take a look at different examples:\\
\begin{itemize}
  \item \textbf{Images} often have multiple thousands of dimensions, but the underlying structure is often far simpler. For example are 28x28 images of handwritten digits 784-dimensional but assuming one-hot encoding the most important information - the digit - it can be explained using 10 dimensions. There are a few more dimensions, stroke width, cursive, position, size. But still far lower than 784. We will take a look at this example later on in \ref{sec:vae}.
  \item \textbf{Speech} or more precisely the recording of it is high-dimensional, while the information it carries (the spoken words) can be represented in a low-dimensional space.
  \item \textbf{...}
\end{itemize}
Given this assumption, building a good representation in the latent space is crucial for minimal reconstruction error as well as good generative modelling (?).
%The VAE framework provides a directed model which aims to infer low-dimensional latent variables $z$ from high-dimensional input data $x$ and reconstruct the input.
Formally, $p(x,z)$ is the joint probability distribution over both input and latent variables while $p(z|x)$ is the conditional probability of the latent variables given input data.
Inferring the posterior distribution $p(z|x)$ is particularly interesting, because it means to enable learning parameters for good latent space representation.
$p(z|x)$ can be expanded using bayes rule to $\frac{p(x|z) p(z)}{p(x)}$.
Computing the nominator is straightforward, $p(z)$ is the probability distribution we chose for the latent space, oftenmost simply a gaussian.
$p(x|z)$ is easy to compute as well, ...? (forward pass?).\\
Meanwhile is the denominator difficult to evaluate, because it requires to consider all possible input combinations.
What we do instead in the variational autoencoders is to derive a lower-bound on $p(z|x)$ using a auxiliary distribution called $q_\theta(z|x)$. The subscript $\theta$ indicates that the distribution is parametrized by an variational term, so that for one $\theta$, $q_\theta(z|x) \approxeq p(z|x)$ holds ($q(z|x)$ doesn't need to depend on $x$).
%Performing statistical inference means computing $p(z|x)$ which is intractable in all but very simple cases (integrating over it, show calc).

>>>>

\section{Variational Autoencoders}
\label{sec:vae}
Variational Autoencoders (VAE) have been introduced in 2014 by Kingma and Welling which uses variational inference to perform efficient inference and learning in deep latent models.
% maybe extra section/sub?

\paragraph{Latent Space}
We assume that the high-dimensional input data $x$ can be explained using low-dimensional latent variables $z$.\\
While this might be surprising at first, most high-dimensional data we encounter can be explained using far fewer dimensions.
Let us take a look at different examples:\\
\begin{itemize}
  \item \textbf{Images} often have multiple thousands of dimensions, but the underlying structure is often far simpler. For example are 28x28 images of handwritten digits 784-dimensional but assuming one-hot encoding the most important information - the digit - it can be explained using 10 dimensions. There are a few more dimensions, stroke width, cursive, position, size. But still far lower than 784. We will take a look at this example later on in \ref{sec:vae}.
  \item \textbf{Speech} or more precisely the recording of it is high-dimensional, while the information it carries (the spoken words) can be represented in a low-dimensional space.
  \item \textbf{...}
\end{itemize}

Given this assumption, building a good representation in the latent space is crucial for minimal reconstruction error as well as good generative modelling (?).
%The VAE framework provides a directed model which aims to infer low-dimensional latent variables $z$ from high-dimensional input data $x$ and reconstruct the input.
\paragraph{Formal Setup}
Formally, $p(x,z)$ is the joint probability distribution over both input and latent variables while $p(z|x)$ is the conditional probability of the latent variables given input data.
Inferring the posterior distribution $p(z|x)$ is particularly interesting, because it means to enable learning parameters for good latent space representation.
$p(z|x)$ can be expanded using bayes rule to $\frac{p(x|z) p(z)}{p(x)}$.
Computing the nominator is straightforward, $p(z)$ is the probability distribution we chose for the latent space, oftenmost simply a gaussian.
$p(x|z)$ is easy to compute as well, ...? (forward pass?).\\
Meanwhile is the denominator difficult to evaluate, because it requires to consider all possible input combinations.
What we do instead in the variational autoencoders is to derive a lower-bound on $p(z|x)$ using a auxiliary distribution called $q_\theta(z|x)$. The subscript $\theta$ indicates that the distribution is parametrized by an variational term, so that for one $\theta$, $q_\theta(z|x) \approxeq p(z|x)$ holds ($q(z|x)$ doesn't need to depend on $x$).
%Performing statistical inference means computing $p(z|x)$ which is intractable in all but very simple cases (integrating over it, show calc).
